---
title: "Assignment 2: Coding Basics"
author: "Isabelle Kuehn"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons/labs in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Canvas.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 55, increasing by fives. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
seq_numbers <- seq(1,55,5)
seq_numbers
#2. 
mean_data <- mean(seq_numbers)
mean_data
median_data <- median(seq_numbers)
median_data

#3. 
determine_stats <- function(mean_data, median_data) {
  if(mean_data > median_data) {
    return("The mean is greater than the median")
  }
  if(median_data > mean_data) {
    return("The median is greater than the mean")
  }
  if(mean_data == median_data) {
    return("The median is equal to the mean")
  }
}
determine_stats(mean_data, median_data)
```

## Basics, Part 2

5.  Create three vectors, each with four components, consisting of (a) student names, (b) test scores, and (c) whether they are on scholarship or not (TRUE or FALSE).

``` {r}
vector_a <- c("Chester", "Isabelle", "Alister", "Elliot")
vector_b <- c(73, 82, 98, 89)
vector_c <- c(FALSE, FALSE, TRUE, TRUE)

```

6.  Label each vector with a comment on what type of vector it is.

```{r}
v_names <- vector_a     #character vector for student names
v_scores <- vector_b   #number vector for test scores
v_schol <- vector_c    #logical vector for scholarship
```

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

```{r}
student_info <- data.frame(v_names, v_scores, v_schol)
```

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
names(student_info) <- c("Student Names", "Test Scores", "Scholorship T/F")
print(student_info)

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: This data frame is different from a matrix because it is used to store different types of data. 

10. Create a function with one input. In this function, use `if`...`else` to evaluate the value of the input: if it is greater than 50, print the word "Pass"; otherwise print the word "Fail". 

11. Create a second function that does the exact same thing as the previous one but uses `ifelse()` instead if `if`...`else `. 

12. Run both functions using the value 52.5 as the input

13. Run both functions using the **vector** of student test scores you created as the input. (Only one will work properly...)

```{r basics 3}
#10. Create a function using if...else

#11. Create a function using ifelse()

#12a. Run the first function with the value 52.5

#12b. Run the second function with the value 52.5

#13a. Run the first function with the vector of test scores

#13b. Run the second function with the vector of test scores
```

14. QUESTION: Which option of `if`...`else` vs. `ifelse` worked? Why? (Hint: search the web for "R vectorization")

> Answer:


**NOTE** Before knitting, you'll need to comment out the call to the function in Q13 that does not work. (A document can't knit if the code it contains causes an error!)