---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Isabelle Kuehn"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2, message = FALSE}
#1
getwd()
library(tidyverse)
library(agricolae)
library(lubridate)
library(here)
#set message = FALSE to have output fit within margins of knitted pdf

lake.raw.data <- read.csv(here(
  "Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), 
  stringsAsFactors = TRUE)
lake.raw.data$sampledate <- as.Date(lake.raw.data$sampledate, 
                                    format = "%m/%d/%y")

#2
default_theme <- theme_classic(base_size = 14) + 
  theme(axis.text = element_text(color = "black"), 
        axis.title.x = element_text(color = "black", size = 12),
        axis.title.y = element_text(color = "black", size = 12),
        plot.title =        element_text(color = "black", size = 13),
        plot.background =   element_rect(fill = "white"),
        panel.background =  element_rect(fill = "aliceblue"),
        panel.grid.major =  element_line(color = "lightgrey"),
        panel.grid.minor =  element_line(color = "grey", linetype = "dotted"),
        legend.text = element_text(color = 'black', size = 9),
        legend.title = element_text(size = 10),
        legend.position = "top")
theme_set(default_theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Lake temperature recorded during July does not change with depth across all lakes.
Ha: Lake temperature recorded during July does change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot, fig.width = 7, fig.height = 4}
#4
July.data <-  lake.raw.data %>%
  filter(month(sampledate) == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()
  

#5
ggplot(July.data, aes(x = depth, y = temperature_C)) +
  geom_point(color = 'slategrey') +
  geom_smooth(method = 'lm', col= 'darkblue') +
  ylim(0,35)+
  xlab("Lake Depth (m)") +
  ylab("Temperature (ºC)") +
  labs(title = "Scatter Plot of Temperature (ºC) Varying by Depth (m)")

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Looking at the resulting figure, the linear trend suggests that temperature decreases as depth increases. The distribution of points along the linear model show that the linear relationship up until around a 7m depth has a strong linear relationship as this is where the concentration of points is highest around the lineear model. The greater the depth, the weaker the linear relationship.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
# remember general setup: y = temp, x = depth -> lm(data ='', y~x)
Temp.Depth.regression <- lm(data = July.data, temperature_C~ depth)
summary(Temp.Depth.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: Looking at the regression results using summary(), the effectiveness of this model can be established. Looking at R-squared value, depth can account for up to 70% of the variability in temperature, which is a strong dependency. To understand the residual standard error, it is best to also look at the degrees of freedom to determine if the model was accurate or not. The residual standard error is under 4, and given that the best results have low residual standard error, this means the data is pretty accurate because there are 9726 degrees of freedom. This means among over 9000 observations, and data points used in the linear regression model, the residual error is ~3.8. The p-value is less than 0.05 which means the results of this model are statistically significant. And to measure change, the coefficient for depth is -1.95 which is the slope of the linear regression model, so for every 1m of depth change, the temperature decreases by 2ºC. Overall, these results show a strong linear relationship between the two variables as depth has a dramatic effect on the temperature.


---

## Multiple Linear Regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
predict_temp_AIC <- lm(data = July.data, temperature_C~year4 + daynum + depth)
step(predict_temp_AIC)
#higher AIC is worse

#10
AIC_improved_regression <- lm(data = July.data, temperature_C~year4 + daynum 
                              + depth)
summary(AIC_improved_regression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: From the AIC results, through the step function as each of the three variables were removed independently, the AIC got larger and the model got worse. This means each variable improved the model, none should be removed, and all are suited to predict temperature. The year, day of data collection, and depth are the final set of explanatory variables that the AIC method suggests should be used to predict temperature. The statistical differences between this model and the previous model are very small. There is a slightly lower RSE indicating a more accurate model and the R-squared value is higher by ~0.1. The p-value is still well below 0.05 so the results remain statistically significant while the degrees of freedom show the observations used in the model changed minimally. This would be a slight improvement ovet the model that only uses depth as the explanatory variable.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
diff_temp_anova <- aov(data = July.data, temperature_C~lakename)
summary(diff_temp_anova)

diff_temp_linear <- lm(data = July.data, temperature_C~lakename)
summary(diff_temp_linear)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Looking at the results of ANOVA test, the p-value is well below 0.05 indicating statistically significant results and showing that the null hypothesis is rejected. This means there is a significant difference in mean temperature among the different lakes. Looking a the linear model, the p-value is the same and confirms that there is significant different between mean temperatures.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, fig.width = 8, fig.height = 5}
#14.
ggplot(July.data, aes(x = depth, y = temperature_C, col=lakename)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = 'lm', se = FALSE, aes(color = lakename)) +
  ylim(0,35)+
  xlab("Lake Depth (m)") +
  ylab("Temperature (ºC)") +
  labs(
    title = 
      "Scatter Plot of Temperature (ºC) Varying by Depth (m) Across Lakes", 
    color = "Lake Name")

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(diff_temp_anova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Looking at the results of the Tukey HSD test, Paul Lake has statistically the same mean temperature as Peter Lake. Between those two there is a difference of -0.498 and a p-value fo 0.22 confirming the null hypothesis that there is no change in mean temperature. Ward Lake has an ever greater statistically equal mean temperature as Peter Lake, seen by the p-value of 0.737 which also confirms the null hypothesis between the two lakes. On the other hand, Central Long Lake has statistically the most distinct mean temperature from the other lakes, seen by continuously low p-values, and differences of over 2 for all lake mean temperatures, but a majority of which are larger than 4. Crampton Lake also has a distinct mean temperature from other lakes, with low p-values but slightly less differences than with Central Long Lake.


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: The HSD Test, seen in code as HSD.test() would group the lake names in pairwise relationships if their mean temperatures were similar. Looking at Peter and Paul Lake, if this test returned these lakes under the same label, it would show they have the same mean temperature. Another option is to run a two-sample t-test. This will test if the mean temperature between Peter and Paul lake is the same, and if it's true, the null hypothesis is confirmed.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
July.data.wrangled <- July.data %>%
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake")
CW_t_test <- t.test(
  July.data.wrangled$temperature_C~July.data.wrangled$lakename)
CW_t_test

#looking at linear model data to confirm
CW_t_test_GLM <- lm(
  July.data.wrangled$temperature_C~July.data.wrangled$lakename)
summary(CW_t_test_GLM)

```

>Answer: Results of the two-sample T-test determines their mean July temperatures are not different with a p-value above 0.05. The difference between the means is around 0.8, and both the low R-squared value and the high p-value suggest the null hypothesis is confirmed. This matches the results from part 16 where the difference was around 0.89 and a p-value of 0.97 showing they are statistically equal means.

